---
title: "main"
author: "haoyi zheng"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

https://huttenhower.sph.harvard.edu/sparsedossa2/

https://github.com/biobakery/biobakery/wiki/SparseDOSSA2
```{r}
rm(list = ls())
devtools::install_github("biobakery/SparseDOSSA2")
```
# Simulating realistic microbial observations with SparseDOSSA2
"Stool" : continuous population structures

```{r}
library(SparseDOSSA2)
library(magrittr)
library(dplyr)
library(ggplot2)
?fit_SparseDOSSA2
? control_fit
# Using the pre-trained dataset for "Stool"
?SparseDOSSA2
?spike
Stool_simulation <- SparseDOSSA2(template = "IBD",  # choose from "Stool", "Vaginal" or "IBD"
                                 new_features = TRUE,  # should new features be simulated
                                 n_sample = 100,  # number of samples to simulate
                                 n_feature = 100,  # number of features to simulate (when 'new_features = TRUE')
                                 verbose = TRUE)  # return detailed info
result = Stool_simulation$simulated_data
#str(Stool_simulation)
```

```{r}
sim <- SparseDOSSA2(template = "Stool", n_sample = 200, new_features = FALSE)
## Using user-provided trained SparseDOSSA2 model:
data("Stool_subset")
fitted <- fit_SparseDOSSA(data = Stool_subset)
sim <- SparseDOSSA2(template = fitted, n_sample = 200, new_features = FALSE)
```


simulated_data
```{r}
dim(Stool_simulation$simulated_data)
head(Stool_simulation$simulated_data)
```


simulated_matrices:
  rel: Relative Abundance
  a_spiked：add spiked-in ;Absolute Abundance
  a_null： null effect ;Absolute Abundance
```{r}
# view all matrices
head(Stool_simulation$simulated_matrices)
# Null (not spiked)
head(Stool_simulation$simulated_matrices$a_null)
# Spiked
head(Stool_simulation$simulated_matrices$a_spiked)
# Relative abundace
head(Stool_simulation$simulated_matrices$rel)



```

params
Returns the parameters used during the simulation. The parameters are provided to the 'template' argument of the 'SparseDOSSA2' function.
```{r}
head(Stool_simulation$params)
```

#  Simulating Realistic Microbial Observations using a User-supplied Template Dataset

```{r}
data("Stool_subset", package = "SparseDOSSA2")
# columns are samples.
Stool_subset[1:2, 1, drop = FALSE]

fitted <- fit_SparseDOSSA2(data = Stool_subset,
                           control = list(verbose = TRUE))
# fitted absence probabilities for first two features in zero-inflated model 
fitted$EM_fit$fit$pi0[1:2]

Stool_subset_simulation <- SparseDOSSA2(template = fitted, 
                                   n_sample = 10, 
                                   new_features = FALSE,
                                   verbose = TRUE)
```

The following is a simple example for simulating longitudinal data with SparseDOSSA 2, using the "Stool" fit.

The spiking will be done with the following characteristics:

Two microbes are spiked, in 50 subjects each with five time points.
The spiked microbes are associated with a case-control variable, and each is associated with its own Gaussian "random effect" variable that's shared across time points within the same subject.

```{r}
feature_up <- 
  "k__Bacteria|p__Proteobacteria|c__Gammaproteobacteria|o__Enterobacteriales|f__Enterobacteriaceae|g__Escherichia|s__Escherichia_coli"
feature_down <- 
  "k__Bacteria|p__Firmicutes|c__Clostridia|o__Clostridiales|f__Ruminococcaceae|g__Faecalibacterium|s__Faecalibacterium_prausnitzii"

n_s <- 10 # 50 subjects
n_t <-  1# 5 time points


```

Now we generate the metadata matrix for spiking that defines which samples are associated with one or more spike scenario.

In our case, is three columns:

The first column is case-control variable.
The other two are subject-specific random effects.

```{r}
mat_metadata <- data.frame(disease = c(rep(0, n_s * n_t/2), rep(1, n_s * n_t /2 )),
                           subject_feature_1 = rep(rnorm(n = n_s), each = n_t),
                           subject_feature_2 = rep(rnorm(n = n_s), each = n_t))
head(mat_metadata)
head(spike_in_setup)
df_metadata_spike <- spike_in_setup %>% tidyr::crossing(associated_property = c("prevalence", "abundance"))
head(df_metadata_spike, n = 20)



set.seed(49)
sim <- SparseDOSSA2::SparseDOSSA2(
    template = "Stool",
    n_sample = nrow(mat_metadata), 
    new_features = FALSE,
    spike_metadata = "both",
    metadata_matrix = as.matrix(mat_metadata), 
    
    verbose = FALSE)

spiked_mat <- sim$simulated_data
spiked_up <- data.frame(sample = colnames(spiked_mat),
                          values = spiked_mat[which(rownames(spiked_mat) == feature_up),],
                          disease = ifelse(mat_metadata$disease == 0, "control", "case"))

p <- ggplot(spiked_up, aes(x = sample, y = values, color = disease))
p <- p + geom_point() + theme_classic() + coord_flip()
p

```



